---
title: "Halloween"
author: "Xiaoyu Lian (A17025943)"
format: pdf
editor: visual
---

## 1. Importing candy data

```{r}
candy_file <- "candy-data.csv"

candy = read.csv(candy_file, row.names=1)
head(candy)
```

> Q1. How many different candy types are in this dataset?

```{r}
nrow(candy)
```

> Q2. How many fruity candy types are in the dataset?

```{r}
sum(candy$fruity)
```

Different Methods

```{r}
candy["Twix", ]$winpercent
candy["Twix", "winpercent"]
```

```{r}
library(dplyr)
candy %>% 
  filter(rownames(candy)=="Twix")%>%
  select(winpercent)
```

```{r}
candy %>% 
  filter(winpercent > 50)|>
  filter(fruity == 1)

top.candy <- candy[candy$winpercent>50,]
top.candy[top.candy$fruity == 1,]
```

## 2. What is your favorate candy? \> Q3. What is your favorite candy in the dataset and what is it’s winpercent value?

```{r}
candy["Almond Joy", ]$winpercent
```

> Q4. What is the winpercent value for “Kit Kat”?

```{r}
candy["Kit Kat", ]$winpercent
```

> Q5. What is the winpercent value for “Tootsie Roll Snack Bars”?

```{r}
candy["Tootsie Roll Snack Bars", ]$winpercent
```

```{r}
library("skimr")
skim(candy)
```

> Q6. Is there any variable/column that looks to be on a different scale to the majority of the other columns in the dataset?

The winpercent has different scale. It has a mean at 50.32, but others are below 1. Remember to scale data before PCA.

> Q7. What do you think a zero and one represent for the candy\$chocolate column?

It means True or false, that whether the candy contains chocolate.

> Q8. Plot a histogram of winpercent values

```{r}
hist(candy$winpercent, breaks = 10)
```

```{r}
library(ggplot2)

ggplot(candy)+ 
  aes(candy$winpercent)+
  geom_histogram(binwidth = 8)+
  theme_bw()
```

> Q9. Is the distribution of winpercent values symmetrical?

No 

>Q10. Is the center of the distribution above or below 50%?

```{r}
summary(candy$winpercent)
```

below 50% \>Q11. On average is chocolate candy higher or lower ranked than fruit candy?

```{r}
chocolate_avg <- mean(candy$winpercent[as.logical(candy$chocolate)], na.rm = TRUE)

fruit_avg <- mean(candy$winpercent[as.logical(candy$fruity)], na.rm = TRUE)

chocolate_avg
fruit_avg
```

> Q12. Is this difference statistically significant?

```{r}
t_test_result <- t.test(candy$winpercent[as.logical(candy$chocolate)],
                        candy$winpercent[as.logical(candy$fruity)])
t_test_result
```

Alternative Way

```{r}
chocolate <- candy |>
  filter(chocolate == 1)
mean(chocolate$winpercent)

fruit <- candy |>
  filter(fruity == 1)
mean(fruit$winpercent)

t_test <- t.test(chocolate$winpercent,fruit$winpercent)
```

## 3. Overall Candy Rankings

`sort` function. If use sort, can sort the value, but lost the identification of the candy

```{r}
play <- c("c","d","a")
sort(play)
order(play)
play[order(play)]
```

> Q13. What are the five least liked candy types in this set?

```{r}
library(dplyr)
candy %>% arrange(winpercent) %>% head(5)
```

```{r}
head(candy[order(candy$winpercent),], n=5)
```

> Q14. What are the top 5 all time favorite candy types out of this set?

```{r}
library(dplyr)
-candy %>% arrange(winpercent) %>% head(5)
```

```{r}
head(candy[order(candy$winpercent, decreasing = T),], n=5)
```

> Q15. Make a first barplot of candy ranking based on winpercent values.

```{r}
library(ggplot2)

ggplot(candy) + 
  aes(winpercent, rownames(candy)) +
  geom_col()
```

> Q16. This is quite ugly, use the reorder() function to get the bars sorted by winpercent?

```{r}
ggplot(candy) + 
  aes(winpercent, reorder(rownames(candy),winpercent))+
  #-winpercent will make the order flip
  geom_col()
```

```{r}
#only tells the chocolate, not enough
#ggplot(candy) + 
  #aes(winpercent, reorder(rownames(candy),winpercent), fill = chocolate)+
  #geom_col()
```

a more customized color scheme that tells chocolate, bars, and fruity etc. all from the one plot

```{r}
#place holder color vector, repeat, for those doesn't belong to the three divisions
my_cols=rep("black", nrow(candy))
#overwrite those 
my_cols[as.logical(candy$chocolate)] = "chocolate"
my_cols[as.logical(candy$bar)] = "brown"
my_cols[as.logical(candy$fruity)] = "pink"
```

```{r}
#use blue for the favorite candy
my_cols[rownames (candy) == "Hershey's Kisses"] <- "blue"
```

```{r}
ggplot(candy) + 
  aes(winpercent, reorder(rownames(candy),winpercent)) +
  geom_col(fill=my_cols) 
```

> Q17. What is the worst ranked chocolate candy?

sixlets

> Q18. What is the best ranked fruity candy?

Starburst

## 4. Taking a look at pricepercent

```{r}
library(ggrepel)

# How about a plot of price vs win
ggplot(candy) +
  aes(winpercent, pricepercent, label=rownames(candy)) +
  geom_point(col=my_cols) + 
  geom_text_repel(col=my_cols, size=3.3, max.overlaps = 8)
#higher value gives more overlap, lower gives less overlap
```

> Q19. Which candy type is the highest ranked in terms of winpercent for the least money - i.e. offers the most bang for your buck?

Reese's Miniature 

> Q20. What are the top 5 most expensive candy types in the dataset and of these which is the least popular?

Nik L Nip, Nestle Smarties, Ring pop, Hershey's Krackel, Hershey's Milk Chocolate
Least popular: Nik L Nip

```{r}
ord <- order(candy$pricepercent, decreasing = TRUE)
top5 <- head( candy[ord,c(11,12)], n=5 )
top5
top5[which.min(top5$winpercent),]
```

> Q21. Make a barplot again with geom_col() this time using pricepercent and then improve this step by step, first ordering the x-axis by value and finally making a so called “dot chat” or “lollipop” chart by swapping geom_col() for geom_point() + geom_segment().

```{r}
ggplot(candy) +
  aes(pricepercent, reorder(rownames(candy), pricepercent)) +
  geom_segment(aes(yend = reorder(rownames(candy), pricepercent), 
                   xend = 0), col="gray40") +
    geom_point()
```

## 5. Exploring the correlation structure

```{r}
library(corrplot)
```

```{r}
cij <- cor(candy)
corrplot(cij, diag = F) #type = "lower"
#type example() in the console can get the example for the function
```

> Q22. Examining this plot what two variables are anti-correlated (i.e. have minus values)?

fruity and chocolate

> Q23. Similarly, what two variables are most positively correlated?

bar and chocolate, chocolate and winpercent

## 6. Principal Component Analysis

```{r}
pca <- prcomp(candy, scale = T)
summary(pca)
```

```{r}
plot(pca$x[,1:2])
```

```{r}
plot(pca$x[,1:2], col=my_cols, pch=16)
```
how to do the original varibales columns contribute to the new PCs. I will look at PC1

```{r}
loadings <- as.data.frame(pca$rotation)
ggplot(loadings)+
  aes(PC1, reorder(rownames(loadings), PC1), fill = PC1)+
  geom_col()
```

```{r}
# Make a new data-frame with our PCA results and candy data
my_data <- cbind(candy, pca$x[,1:3])
```

```{r}
p <- ggplot(my_data) + 
        aes(x=PC1, y=PC2, 
            size=winpercent/100,  
            text=rownames(my_data),
            label=rownames(my_data)) +
        geom_point(col=my_cols)
p
```

```{r}
library(ggrepel)

p + geom_text_repel(size=3.3, col=my_cols, max.overlaps = 8)  + 
  theme(legend.position = "none") +
  labs(title="Halloween Candy PCA Space",
       subtitle="Colored by type: chocolate bar (dark brown), chocolate other (light brown), fruity (red), other (black)",
       caption="Data from 538")
```

```{r}
#library(plotly)
#ggplotly(p)
```

```{r}
par(mar=c(8,4,2,2))
barplot(pca$rotation[,1], las=2, ylab="PC1 Contribution")
```

> Q24. What original variables are picked up strongly by PC1 in the positive direction? Do these make sense to you?

The fruity, hard, and pluribus are picked up strongly.Yes, because they are correlated to each others
